---
title: "Logistic Regression Analysis"
author: "Yalin Yang"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
  word_document:
    toc: no
    toc_depth: '3'
--- 

# Logits to Probability

Logits $\in [+6,-6]$  and Regression line as $y = \beta_0 + \beta_1 * Logits$

```{r}
## linear form in x
b0 <- 1   # Play with different parameters to see their effects on the Probs
# b0 <- 0
b1 <- 5

x <- seq((-6-b0)/b1,(6-b0)/b1,length.out=200)  # Adjusted scale to b0 and b1
#x <- seq((-40-b0),(40-b0),length.out=200)  # Adjusted scale to b0 and b1
L <- b0+b1*x
```

## Different forms for p and (1-p)

```{r}
## Probs of first category
p1 <- 1/(1+exp(-L))
p2 <- exp(L)/(1+exp(L))

## Probs of second category
np1 <- 1/(1+exp(L))
np2 <- exp(-L)/(1+exp(-L))

## Logistic curve turning point
medX <- -b0/b1
```

## Plot for different functional forms

```{r fig.height=8, fig.width=12}
layout(matrix(1:2,nrow=1,ncol=2))
  ## Prob(Success)
  plot(x,p1,ylab="1/(1+exp(-L))",ylim=c(0,1),type="l", lwd=2, col="red",main="Pr(X < x)")
  abline(h=c(0,0.5,1),lty=2); abline(v=medX,lty=5)
  ## Prob(Failure)
  plot(x,np1,ylab="1/(1+exp(L))",ylim=c(0,1),type="l", lwd=2,col="blue", main="1 - Pr(X < x)")
  abline(h=c(0,0.5,1),lty=2); abline(v=medX,lty=5)
layout(matrix(1,nrow=1,ncol=1))

```

## Check equality of logit against L

```{r}
all.equal(L,log(p1/np1))
```

# Logistic Regression Analysis [Binary output]

## Quick Look of dataset

```{r}
library(car)
CloseVote <- foreign::read.spss("G:\\UTD_Classes\\2020Spring\\GISC7310_AdvancedDataAnalysis\\07Logistic Regression Analysis\\SchoolClosing.sav", to.data.frame=TRUE)

## Evaluate which variables are factors
sapply(CloseVote,is.factor)

```

Exploratory plots: barwidth _proportional_ to the NumOfObs in interval

```{r fig.height=10, fig.width=15}
par(mfrow = c(2,2))
plot(close~lived, data=CloseVote)
plot(close~educ, data=CloseVote)
plot(close~contam, data=CloseVote)
plot(close~hsc, data=CloseVote)
plot(close~female, data=CloseVote)
plot(close~kids, data=CloseVote)
plot(close~nodad, data=CloseVote)
```

## Logistic modeling

Just intercept an intercept model

```{r}
## generalize linear model
GLM.00 <- glm(close ~ 1, family=binomial(logit), trace=TRUE, data=CloseVote) # just intersecpt
summary(GLM.00)
cat("Deviance: ", logLik(GLM.00)*-2)
```

Tansfer intercept from logis to probability model

```{r}
1/(1+exp(-(coef(GLM.00)[1])))            # predicted prob in favor of closing 1/(1+exp(-b_0)) 
mean(unclass(CloseVote$close)-1)         # same as average zeros and ones
```

Bi-variate model "lived" and intercept

```{r}
GLM.01 <- glm(close ~ lived, family=binomial(logit), trace=TRUE, data=CloseVote)
summary(GLM.01)  #slope is for logit, not for probability
cat("Deviance: ", -2*logLik(GLM.01))
```

## Likelihood Ratio Test

The "hard" way

```{r}
( LR <- -2*(logLik(GLM.00)-logLik(GLM.01)) )
( pchisq(LR[1], df=1, lower.tail=F) )
```

The "easy" way

```{r}
anova(GLM.00,GLM.01,test="LRT")
```

## Alternative: The probit model 

```{r}
GLM.probit <- glm(close ~ lived, family=binomial(link=probit), trace=TRUE, data=CloseVote)
summary(GLM.probit)
logLik(GLM.probit)
```

## Full model and Restricted model

Full model with all variables and interaction term nodad

```{r fig.height=8, fig.width=12}
GLM.02 <- glm(close ~ lived + educ + contam + hsc + nodad + female + kids ,
              family=binomial(logit), trace=TRUE, data=CloseVote)
summary(GLM.02, correlation=F)
vif(GLM.02)
logLik(GLM.02)
# confint(GLM.02, level=0.95, type="Wald")
```

Restricted model without female and kids

```{r}
GLM.03 <- glm(close ~ lived + educ + contam + hsc + nodad ,
             family=binomial(logit), data=CloseVote,
             control=list(epsilon=1e-15,maxit=50, trace=TRUE))
summary(GLM.03, correlation=TRUE)
logLik(GLM.03)
```

### Likelihood Ratio Test

```{r}
anova(GLM.03,GLM.02,test="LRT")
```

## Effect plots

```{r fig.height=8, fig.width=12}
library(effects)
## all independent variables with the others at average. 
## Note type="response" for probability scale
plot(allEffects(GLM.03), type="response", ylim=c(0,1), ask=FALSE)
```

```{r}
## Group specifice effect plots
summary(CloseVote)
```

### Low prob respondent
```{r fig.height=8, fig.width=12, message=FALSE, warning=FALSE}
eff.GLM.low <- effect("lived",GLM.03,
                      given.values=c(educ=20,"contamyes"=0,"hscyes"=0,"nodadyes"=1))
plot(eff.GLM.low, type="response", ylim=c(0,1), ylab=expression(Pr(Y[i]=="Close")), 
     main="Low Probability Respondents")
```

### Average prob respondent

```{r fig.height=8, fig.width=12, message=FALSE, warning=FALSE}
eff.GLM.average <- effect("lived",GLM.03)
plot(eff.GLM.average, ylim=c(0,1), type="response", ylab=expression(Pr(Y[i]=="Close")),    # ylim is in terms of probs
     main="Average Probability Respondents")
```

### High prob respondent

```{r fig.height=8, fig.width=12, message=FALSE, warning=FALSE}
eff.GLM.hi <- effect("lived",GLM.03,
                     given.values=c(educ=6,"contamyes"=1,"hscyes"=1,"nodadyes"=0))
plot(eff.GLM.hi, type="response", ylim=c(0,1), ylab=expression(Pr(Y[i]=="Close")), 
     main="High Probability Respondents")
```

# Residual Exploration

## Response Residuals

```{r fig.height=8, fig.width=12}
resid.GLM.03 <- residuals(GLM.03, type="response")   # check ?residuals.glm
pred.GLM.03 <- predict(GLM.03, type="response")      # check ?predict.glm
plot(pred.GLM.03, resid.GLM.03, ylim=c(-1,1), xlim=c(0,1), 
     ylab="Response Residuals", xlab="Predicted Probabilities")
abline(h=0, lty=5)
lines(lowess(pred.GLM.03, resid.GLM.03),lwd=2)  # Smoothed function to see the residual behavior
```


## Pearson Residuals (standardized residuals)

```{r fig.height=8, fig.width=12}
resid.GLM.03 <- residuals(GLM.03, type="pearson")   
pred.GLM.03 <- predict(GLM.03, type="response")      
plot(pred.GLM.03, resid.GLM.03, xlim=c(0,1), 
     ylab="Pearson Residuals", xlab="Predicted Probabilities")
abline(h=0, lty=5)
lines(lowess(pred.GLM.03, resid.GLM.03),lwd=2)  # Smoothed function to see the residual behavior
```

# Logistic Regression Analysis [Rates]

Weighted GLM specification

```{r message=FALSE, warning=FALSE}
insects <- data.frame(popDensity=c(1,4,10,22,55,121,210,444),
                      females=c(1,3,7,18,22,41,52,79),
                      males=c(0,1,3,4,33,80,158,365))

insects$totalPop <- insects$females+insects$males
insects$rateMale <- insects$males/insects$totalPop
glm.weight <- glm(rateMale~log(popDensity), weights=totalPop, data=insects, family=binomial)
summary(glm.weight)
```

```{r fig.height=8, fig.width=12}
## Plot prediction
xAxis <- seq(0,7,by=0.001)                                                    # value range of log(popDensity)
probPred <- predict(glm.weight,list(popDensity=exp(xAxis)), type="response")   # type="response" give predicted probabilities
plot(log(insects$popDensity), insects$rateMale, ylim=c(0,1), xlim=c(0,7), 
     ylab=expression(Pr(Y[i] =="Male")),xlab=expression(log("Population Density")), 
     pch=16, col="blue")
lines(xAxis,probPred, col="red")
abline(h=c(0,1),lty=3)
```

