---
title: "Regression Diagnostics"
author: "Yalin Yang"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
--- 

# Standard Regression Diagnositcs

## Quick view of dataset

Evaluate monthly cycle and variance heterogeneity

```{r fig.height=8, fig.width=12}
sandusky <- foreign::read.spss("SanduskyTemperature.sav", use.value.labels = TRUE, to.data.frame = TRUE)
boxplot(avg7447~month, data=sandusky, ylab="Average Temperature in Fahrenheit", main="Sandusky Climate 1990-1999")
```

```{r fig.height=8, fig.width=12}
plot(avg7447~time.idx, data=sandusky, main="Monthly Temperature Variation at Sandusky, Ohio, from 1990 to 1999",
     xlab="Sequence in Months", ylab="Temperature", type="l")
```

## Generate harmonic variables and add them to the data-frame

Fouier regression with 2 wave parameters

```{r}
sandusky$r.cos <- cos(sandusky$time.idx/12*2*pi)
sandusky$r.sin <- sin(sandusky$time.idx/12*2*pi)
fourier1.lm <- lm(avg7447~time.idx+r.cos+r.sin, data=sandusky)
summary(fourier1.lm,cor=T)
```

Variance inflation factors (reported in variance NOT std)

```{r message=FALSE, warning=FALSE}
library(car)
vif(fourier1.lm)
```

covariance among estimated parameters

```{r}
round(vcov(fourier1.lm),2) 
```

## Fixed effect panel model

```{r}
month.lm <- lm(avg7447~time.idx + month, data=sandusky)
summary(month.lm)
```

## Diagnostic plots

### Test of normality based on t-distribution

```{r fig.height=8, fig.width=12, message=FALSE, warning=FALSE}
qqPlot(fourier1.lm) 
```

### Partial effects plots

```{r fig.height=8, fig.width=12}
avPlots(fourier1.lm) 
```

### Residual plots (Tukey test)

```{r fig.height=8, fig.width=12}
residualPlots(fourier1.lm)
```

## Update the model by adding I(time.idx^2)

```{r}
fourier2.lm <- update(fourier1.lm, .~.+I(time.idx^2))
summary(fourier2.lm)
```

recheck for non-linearity

```{r}
anova(fourier1.lm, fourier2.lm)
```

```{r fig.height=8, fig.width=12}
residualPlots(fourier2.lm) 
```

## Get residuals

```{r}
(resid <- residuals(fourier2.lm))[1:10]
(std.resid <- rstandard(fourier2.lm))[1:10]
(student.resid <- rstudent(fourier2.lm))[1:10]
```

## Other diagnositic measures

### DFBeta

```{r fig.height=8, fig.width=12}
dfbeta.values <- dfbetas(fourier2.lm)
boxplot(dfbeta.values, main="dfBetas values")
```

### Cook distance

```{r fig.height=8, fig.width=12}
cook.values <- cooks.distance(fourier2.lm)
boxplot(cook.values, main="cook-values")
# boxplot(cook.values, main="cook-values", id.n=2)
```

### Leverage Plot

```{r fig.height=8, fig.width=12}
leverage.values <- hatvalues(fourier2.lm)
boxplot(leverage.values, main="leverage values")
```

```{r fig.height=8, fig.width=12}
car::influencePlot(fourier2.lm)
```

**Be careful: inspect scale of Bonferroni p-values**

```{r fig.height=8, fig.width=12}
car::influenceIndexPlot(fourier2.lm)
```

