---
title: "The Generalized Linear Model"
author: "Yalin Yang"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
--- 

# Poisson Distribution for Probability [0-1]

## Two methods for pushing backward to probability from logits.

Explore the shape of the **logistic probability curve** in dependence of the $logit = a + b * x$, 
The first method using equation as $prob = \frac{1}{1 + exp(-logit)} $

```{r fig.height=8, fig.width=12}
x <- seq(-9,9,by=0.1)

a <- 1      #Intercept
b <- 0.5   # slope

logit <- a+b*x
prob <- 1/(1+exp(-(logit)))
```

There is no big difference when you use scale variable to transfer logit($prob = \frac{logit}{sqrt(\pi^\frac{2}{3})} $) to a normal distribution

```{r fig.height=8, fig.width=12}
plot(x,prob,type="l",ylim=c(0,1),col="red",lwd=2,
     ylab="Probability")

abline(v=-a/b,lty=3)            # 0.5 prob point is at -a/b
abline(h=0.5,lty=3)
##  Overlay probit curve
p2lScale <- sqrt(pi^2/3)                 # Scale to match variance of logit and probit functions
prob <- pnorm(logit/p2lScale)            # This is the probit
lines(x,prob,col="blue",lwd=2)
abline(h=c(0,1),lty=2)
legend(2,0.9, legend=c("logit","probit"), lty=1, col=c("red","blue"),bty="n")
```

## Simulate binary data and estimate model

```{r}
rm(list=ls(all=TRUE))
a <- 1; b <- 1                              # Increase slope to see increasing discrimination power
nObs <- 50                                  # Reduce number of observations to see instability of results
#nSize <- rep(1,nObs)                       # Binary distribution
nSize <- rpois(nObs,lambda=20)+1            # Binomial distribution's population at risk
xObs <- runif(nObs,-5,5)                    # Random X
yObs <- vector(mode="numeric", length=nObs) # Initialize vector for Probs

## Generate random 0/1 or rates with given probability and population at risk
yProb <- 1/(1+exp(-(a+b*xObs)))             # Get probabilities by inverse logit in linear predictor
for (i in 1:nObs) {                         # Simulate binomial distributed counts
  yObs[i] <- rbinom(1, nSize[i], yProb[i])
  }
(simDF <- data.frame(yObs, yProb, nSize, xObs))[1:10,]
```

Fit model **binary and binomial model**. Binary shortcut: glm(yObs~xObs, ...)

```{r}
rate <- yObs/nSize
sim.GLM1 <- glm( rate ~ xObs, weights=nSize,     # Rates specification with 
               family=binomial(link="logit"),    # population at risk weights 
               data=simDF)     
summary(sim.GLM1)

```

Plot results

```{r fig.height=8, fig.width=12}
plot(xObs, predict(sim.GLM1, type="response"), ylim=c(0,1), type="n")
if (sum(nSize - rep(1,nObs))==0) {             # For binary model
    rug(xObs[yObs==0],side=1)
    rug(xObs[yObs==1],side=3)
  } else {                                     # For binomial model with rates
    points(xObs,yObs/nSize, pch=20) }

## Add predicted line
lines( seq(-5,5,by=0.1),
       predict(sim.GLM1,newdata=data.frame(xObs=seq(-5,5,by=0.1)),type="response"))
```


## Explore shapes of the Poisson distribution

```{r fig.height=10, fig.width=15}
plot(0:40,dpois(0:40,1), xlab="Count", ylab="Density", ylim=c(0,0.5), type="n",
     main="Shape of Poisson Distributions")
for (i in c(1,2,5,10,20)) lines(spline(0:40,dpois(0:40,i)))
text(1, 0.37, expression(lambda,"  =1"),pos=4)
text(2.4, 0.26, expression(lambda,"  =2"),pos=4)
text(4, 0.20, expression(lambda,"  =5"),pos=4)
text(6.8, 0.14, expression(lambda,"  =10"),pos=4)
text(17, 0.11, expression(lambda,"  =20"),pos=4)
```

# Poisson Distribution for Counts

## Initialize data

```{r}
nObs <- 50
a <- 1; b <- 1
xLow <- 0; xHi <- 3
# Generate Data
xObs <- runif(nObs,xLow,xHi)
lambda <- exp(a+b*xObs)           # inverse link function "exp" for Poisson GLM
yObs <- rep(0,nObs)               # initialize output vector
for (i in 1:nObs) {               # simulate counts subject to expectation lambda
  yObs[i] <- rpois(1,lambda[i])
  }
(simDF <- data.frame(yObs,lambda,xObs))[1:10,]
```

## Estimate Poisson model

```{r}
sim.GLM <- glm(yObs~xObs, family=poisson(link="log"), data=simDF)
summary(sim.GLM)
```

**Plot model**

```{r fig.height=8, fig.width=12}
plot(xObs, predict(sim.GLM,type="response"), ylim=c(0,max(simDF$yObs)), type="n",
     main="Poisson regression")
points(xObs,yObs,pch=20)
lines( seq(xLow,xHi,by=0.1),
       predict(sim.GLM,newdata=data.frame(xObs=seq(xLow,xHi,by=0.1)),type="response"),
       lwd=2)
```

# Poisson model for counts with overdispersion

## Initialize data

```{r}
addSpread <- runif(nObs, 0, sqrt(lambda))
yObs[yObs < lambda] <- yObs[yObs < lambda] - addSpread[yObs < lambda]
yObs[yObs > lambda] <- yObs[yObs > lambda] + addSpread[yObs > lambda]
yObs[yObs < 0] <- 0
(simDF <- data.frame(yObs,lambda,xObs,addSpread))[1:10,]
```

## Estimate Poisson model with offset

```{r}
sim.GLM <- glm(yObs~xObs, family=quasipoisson, data=simDF)  # log of offset important
summary(sim.GLM)
```

**Plot Model**

```{r fig.height=8, fig.width=12}
plot(xObs, predict(sim.GLM,type="response"), ylim=c(0,max(simDF$yObs)), type="n",
     main="Poisson regression with overdispersion")
points(xObs,yObs,pch=20, col="red")
lines( seq(xLow,xHi,by=0.1),
       predict(sim.GLM,newdata=data.frame(xObs=seq(xLow,xHi,by=0.1)),type="response"),
       col="red")
```

# Poisson model for counts with Offset

## Initialize data

```{r}
rm(list=ls(all=TRUE))
nObs <- 50
a <- 1; b <- 1
xLow <- 0; xHi <- 3

# Generate Data
xObs <- runif(nObs,xLow,xHi)
shift <- rep(5,nObs)                # Shifts lambda by factor 5
lambda <- exp(a+b*xObs+log(shift))  # the Offset is on the same scale as expected counts
yObs <- rep(0,nObs)
for (i in 1:nObs) {
  yObs[i] <- rpois(1,lambda[i])
  }
(simDF <- data.frame(yObs,lambda,xObs,shift))[1:10,]
```

## Estimate Poisson model with offset

```{r}
sim.GLM <- glm(yObs~xObs, family=poisson(link="log"), offset=log(shift), data=simDF)  # log of offset important
summary(sim.GLM)
```

**Plot Model**

```{r fig.height=8, fig.width=12}
plot(xObs, predict(sim.GLM,type="response"), ylim=c(0,max(simDF$yObs)), type="n",
     main="Poisson model with fixed offset")
points(xObs,yObs,pch=20, col="blue")
lines(seq(xLow, xHi, by=0.1), predict(sim.GLM, type="response",
                              newdata=data.frame(xObs=seq(xLow, xHi, by=0.1), shift=2)), col="red")
```

# Migration Model

## Initialize data

```{r}
setwd("G:\\UTD_Classes\\2020Spring\\GISC7310_AdvancedDataAnalysis\\08The Generalized Linear Model")
berlin <- foreign::read.spss("bmigvec91_92.sav", use.value.labels=TRUE, to.data.frame=TRUE)
str(berlin)
```

**index i for origin, index j for destination, and index ij for inter-regional links**
```{r}
attach(berlin)
i <- org
j <- dest

lnpi91 <- log(poporg91)
lnpj91 <- log(popdest91)

lnpi92 <- log(poporg92)
lnpj92 <- log(popdest92)

lnmij91 <- mij91
lnmij91[mij91 >0 ] <- log(mij91[mij91 >0 ])

lndij <- dij                         # Make sure that the log(dij=0) is treated properly
lndij[dij > 0] <- log(dij[dij > 0])  # dij = 0 remains zero
```

## Plain Migration Model [Gravity Model]

```{r}
mod01 <- glm(mij92 ~ lnpi92+lnpj92+lndij, weights = cwt, data=berlin, family = poisson(log)) 
summary(mod01)
```

**Check Disperion**

```{r}
mod01b <- glm(mij92 ~ lnpi92+lnpj92+lndij, weights = cwt, data=berlin, family =quasipoisson)
summary(mod01b)
```

## Plain Migration Model with Offset
variation around the migration flow in 1991

```{r}
mod02 <- glm(mij92 ~ lnpi92+lnpj92+lndij, offset=lnmij91, weights = cwt, data=berlin, family = poisson(log))
summary(mod02)
```

## Origin specific distance decay parameter

There is a correlation between origins and the migration distance, so we need to applied a nested model

```{r fig.height=8, fig.width=12}
bymedian <- with(berlin, reorder(i, dij, median))
boxplot(dij ~ bymedian, subset=(i!=j), data=berlin,
        main="Distance Distribution from Each District")
```

```{r}
mod03 <- glm(mij92 ~ lnpi92+lnpj92+ lndij+lndij:i, weights = cwt, data=berlin, family = poisson, contrast = list(i="contr.sum"))
summary.glm(mod03)
```

The distance decay paramter

```{r}
coef23 <- -sum(coef(mod03)[5:26])   # Get missing last distance decay parameter
data.frame(orgDist <- c(coef(mod03)[5:26],coef23))
```

