---
title: "Spatial Autocorrelation & Heteroscedasticity"
author: "Yalin Yang"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
--- 

# Maximum Likelihood

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, prompt = TRUE, comment = "R>", cache = FALSE)
```

## Evaluate the likelihood and log-likelhood functions

The following code-chunks evaluate the likelihood function 
$L(\pi \mid x,n)=\pi^x \cdot (1-\pi)^{n-x}$ and the log-likelihood function of a binary distributed outcomes $x_i \in \{0,1\}$. 

Set up the model:
```{r def}
pi <- 1:99/100                                 # sequence of valid pi values
x <- 7                                         # number of successes
n <- 10                                        # number of trails
likFunc <- function(x,n,pi) pi^x*(1-pi)^(n-x)  # likelihood function
```

It shows, by example, that the maximum $\pi_{max}$ of both functions with the number of successes `r x` and the number of trails `r n` is the *identical*.

Plot the likelihood function $L(\pi \mid x,n)$

```{r lik, fig.height=8, fig.width=12}
plot(pi,likFunc(x,n,pi), type="l", xlab=expression(pi), 
     ylab="likelihood function")
abline(v=x/n, lty=5, col="red")
```

Plot the log-likelihood function $\log(L(\pi \mid x,n))$

```{r loglik, fig.height=8, fig.width=12}
plot(pi,log(likFunc(x,n,pi)), type="l", xlab=expression(pi), 
     ylab="log-likelihood function")
abline(v=x/n, lty=5, col="red")
```

The numerical solution for $\hat{\pi} = \frac{x}{n}$
The log-likelihood function is:
$$
\log(L(\pi)) \stackrel{\rm def}{=}
    \log(\pi^x \cdot (1-\pi)^{n-x}) = x \cdot \log \pi + (n-x) \cdot \log (1-\pi)  
$$

Its first derivative becomes:
$$
\frac{d \log(L(\pi))}{d \pi}  =  \frac{x}{\pi} + \frac{n-x}{1-\pi} \cdot -1 =  \frac{x}{\pi} - \frac{n-x}{1-\pi}
$$

Setting the first derivative equal to zero gives the desired estimator \hat{\pi} = \frac{x}{n}:
$$
\frac{x}{\pi} - \frac{n-x}{1-\pi} = 0 \Rightarrow \hat{\pi} = \frac{x}{n}
$$

# Heteroscedasticity

## Read data & explore data

```{r fig.height=8, fig.width=12, message=FALSE, warning=FALSE}
rm(list=ls())
library(TexMix)
setwd("G:\\UTD_Classes\\2020Spring\\GISC7310_AdvancedDataAnalysis\\06Spatial Autocorrelation")
Bladder <- foreign::read.spss("bladder_wmp2.sav",to.data.frame=TRUE)

car::scatterplotMatrix(~bladp2+lungp1+popden, data=Bladder)
car::scatterplotMatrix(~log(bladp2)+log(lungp1)+log(popden), data=Bladder)
hist(Bladder$pop, breaks=20, freq=FALSE, xlab="Population at Risk")
lines(density(Bladder$pop), col="red")
```
Note: these data are _areal aggregates_ therefore there will be some _ecological biases_ in the model estimates.

## Variables description
* The dependent variable `bladp2` measure the age-adjusted male bladder cancer mortality rate per 100.000 population at risk for the 508 US State Economic Areas (SEA) for the period of 1970 to 1994.

* The independent variable `lungp1` is the male lung cancer rate for the period 1950 to 1969. It is a proxy variable for smoking. Smoking is both a risk factor for lung and bladder cancer. Lung cancer has a shorter latency period than bladder cancer. Therefore, the 1950 to 1969 period is used.

* The independent variable `popden` is a proxy variable for behavioral and environmental factors. The populations behavior as well as environmental conditions are different in urban, suburban and rural SEAs.

* The weights variable `pop` is the count of the population at risk in each SEA for the period 1970 to 1994. It varies substantially from SEA to SEA. Thus the SEA's rates of `bladp2` have different denominators and are expected to have varying variances, which cause heteroscedasticity.

## Basic model and heteroscedasticity test

```{r}
lmBase <- lm(log(bladp2)~log(lungp1)+log(popden), data=Bladder)
summary(lmBase)

car::ncvTest(lmBase, data=Bladder) 
car::ncvTest(lmBase, var.formula=~log(pop), data=Bladder)
```

Concurring with epidemiological knowledge, the higher the smoking rate (i.e., `lungp1`) the higher the bladder cancer rate. More densely populated SEA's have are populated the higher the bladder cancer rate. 

The Breusch-Pagan test `car::ncvTest` preforms a score test of the hypothesis of constant error variance against the alternatives [a] that the error variance changes with the level of the fitted values (first test), or [b] with a linear combination of predictors (second test). Notice that the weights variable in the second test is entered log-transformed.

* The unspecific test without an explicit heteroscedasticity model is borderline significant. 

* In contrast, assuming that the error variance is a function of the underlying population at risk is highly significant.

## Underlying idea of model calibration
See Kleiber & Zeileis pp 76-78 "Weighted Least Squares".  At the first step, use the model 

$$
 \log{\hat{\sigma}_i^2}={\tt fitted}(\log{\epsilon_i^2} \sim \log pop_i) 
$$

to predict the heteroscedastic error variance. In a subsequent step use the inverse variance $w_i=\frac{1}{\exp(\hat{\sigma}_i^2)}$ as weight in an updated regression $y \sim x_1+x_2+ \cdots +x_k, {\tt weight}=w_i$. Note: This way observations with higher uncertainty are down-weighted. It is possible to iterate.

```{r}
auxreg <- lm(log(residuals(lmBase)^2)~log(Bladder$pop)) 
summary(auxreg)
```

```{r fig.height=8, fig.width=12}
plot(log(residuals(lmBase)^2)~log(Bladder$pop)); abline(auxreg, col="red")
title("Heteroscedastic lm-Residuals")
```

```{r}
## Weighted Regression
predLogSigma2 <- fitted(auxreg)
lmUpdated <- update(lmBase, weights=1/exp(predLogSigma2))
summary(lmUpdated)
```

## Use `lmHetero()` to model Heteroscedasticity
To account for heteroscedasticity related to `log(pop)` the `lmHetero` function in the package `TexMix` can be used. See its online help on how to call the function and what objects it returns.

```{r}
lmH1 <- lmHetero(lnbladd~lnlung1+lnpopden | log(pop), data=Bladder)
summary(lmH1)                         # coefficients of lmH1 and lmBase are similar
# names(lmH1)
```

Compared to the OLS model the estimated regression coefficients have not changed substantially because even under heteroscedasticity the OLS coefficients are unbiased. However, their standard errors will change.

The `gamma coefficient` of `log(pop)` is _negative_ and significant. Thus with an increasing population at risk $n_i$, as suggested by theory, the error variance $\sigma_i^2$ shrinks:
$$
\sigma_i^2 \sim \frac{1}{n_i}
$$

The likelihood ratio test indicates that, compared to the log-likelihood of the plain OLS model, the log-likelihood of the adjusted model is _significantly_ larger.

The `lmHetero( )` returns also a vector of case weights, which can be use in the standard `lm( )` function. This allows to perform the standard model diagnostics of the OLS model. The estimated regression coefficients are identical to those of the `lmHetero` model. The residuals obtained with the function `weighted.residuals( )` are free of heteroscedasticity with respect to the weights variable.

```{r}
## Once weights are estimated, diagnositcs could proceed with lm() and the weights-option
wlm1 <- lm(log(bladp2)~log(lungp1)+log(popden), data=Bladder, weights=lmH1$weights)
summary(wlm1)                         # Results of lmH1 and wlm1 are identical
wResid <- weighted.residuals(wlm1)    # Function to adjust residuals for heteroscedasticity
```

The example shows how a linear combination of several weights variables can be used.

```{r}
##
## Model with quadratic weights structure
##
lmH2 <- lmHetero(log(bladp2)~log(lungp1)+log(popden) | log(pop)+I(log(pop)^2), data=Bladder)
summary(lmH2)

##
## Manual LR-Test: Quadratic versus linear weights structue
##
(chi <- -2*(lmH1$logLikeH1-lmH2$logLikeH1))
(pchisq(chi,df=1, lower.tail=F))
```

The manual likelihood ratio test compares the model with `| log(pop)` against the augmented model `| log(pop)+I(log(pop)^2`. In this case, there is no significant improvement using the augmented model and its `Gamma Coefficients` are no longer significant.

# Spatial Autocorrelation

## Import data

```{r fig.height=8, fig.width=12, message=FALSE, warning=FALSE}
rm(list=ls(all=TRUE))
library(car); library(maptools); library(spdep); library(TexMix)
##
## Read Poly Shapefiles (readShapePoly in library maptools)
##
setwd('./Italy')
getinfo.shape("Provinces.shp")
neig.shp <- rgdal::readOGR(dsn=getwd(), layer="Neighbors", integer64="warn.loss")
prov.shp <- rgdal::readOGR(dsn=getwd(), layer="Provinces", integer64="warn.loss")
summary(prov.shp)

proj4string(prov.shp)                                    # map projection
prov.centroid <- coordinates(prov.shp)                   # Get province centroids
prov.bbox <- bbox(prov.shp)                              # province bounding box for map region
```

## Bipolar Map

Qualitative map theme of `REGION`

```{r fig.height=8, fig.width=12}
table(prov.shp$REGION)

plot(neig.shp,axes=T,col=grey(0.9), border="white",      # background: neighboring countries
     xlim=prov.bbox[1,], ylim=prov.bbox[2,])                     
mapColorQual(prov.shp$REGION, prov.shp, map.title="Italy's Regions",
                          legend.title="Region", add.to.map=T)   
```

Gradient map theme of `TOTFERTRAT`

```{r fig.height=8, fig.width=12}
plot(neig.shp,axes=T,col=grey(0.9),border="white",       # first background (axes=T adds lat/long frame)
     xlim=prov.bbox[1,],ylim=prov.bbox[2,])              # within bounding box
mapColorRamp(prov.shp$TOTFERTRAT, prov.shp, breaks=8,    # second add map
              map.title="Spatial Pattern of Fertility Rate",
              legend.title="Fertility Rate", 
              legend.cex=1.7, add.to.map=T)
```

## Basic regression model

```{r fig.height=8, fig.width=12}
## Regression model for fertility including model diagnostics
scatterplotMatrix(~TOTFERTRAT+ILLITERRAT+FEMMARAGE+DIVORCERAT+TELEPERFAM, data=prov.shp,
                   smooth=list(span = 0.35, lty.smooth=1, col.smooth="red", col.var="salmon"),
                   regLine=list(col="green"))

fert.lm <- lm(TOTFERTRAT~ILLITERRAT+FEMMARAGE+DIVORCERAT+TELEPERFAM, data=prov.shp)
summary(fert.lm,corr=F)
vif(fert.lm)

# Perform Residual Diagnostics
influenceIndexPlot(fert.lm, id=list(n=3,labels=prov.shp$PROVNAME))
fertResid <- residuals(fert.lm)
```

## Identify potential outlier

```{r fig.height=8, fig.width=12}
## Why is Bolzano-Bozen an extreme observation? Shall we delete it?
idx.max <- which.max(abs(fertResid))        # Get index of a record with "outlying" observation 

## Map potential outlier
extremeObs <- rep(0, length(fertResid))
extremeObs[idx.max] <- 1
extremeObs <- factor(extremeObs, labels=c("Population","Outlier"))
table(extremeObs)
plot(neig.shp,axes=T,col=grey(0.9),border="white",                 # background: neighboring countries
     xlim=prov.bbox[1,],ylim=prov.bbox[2,])                     
mapColorQual(extremeObs, prov.shp, map.title="Potential Outlier",
              legend.title="Outliers", add.to.map=T)        

## Inspect outlier
prov.shp@data[idx.max,]                      # List info of record centre is outlier

## Delete outlier or update information
#prov.shp <- prov.shp[-idx.max ]             # Delete extreme observation from shapefile 
prov.shp@data[idx.max, "TOTFERTRAT"] <- 1.2  # Or change its value within the shapefile

## Continue with updated dataset
fert.lm <- lm(TOTFERTRAT~ILLITERRAT+FEMMARAGE+DIVORCERAT+TELEPERFAM,data=prov.shp) # update model
summary(fert.lm)
influenceIndexPlot(fert.lm, id=list(n=3, labels=prov.shp$PROVNAME))
```

## Check for heteroscedasticity

```{r}
## Check for heteroscedasticity
fert.fgls <- lmHetero(TOTFERTRAT~ILLITERRAT+FEMMARAGE+DIVORCERAT+TELEPERFAM | log(FEMPOP94),
                      data=prov.shp)
summary(fert.fgls)
```

Note the possitive sign of $\gamma_2$ is contradicting theory. This is marginal significant of the heteroscedastic model and will be used subsequently for demonstration purposes.

Therefore, the lm-model is updated to a weighted lm-model.
```{r wlm}
fert.wlm <- lm(TOTFERTRAT~ILLITERRAT+FEMMARAGE+DIVORCERAT+TELEPERFAM, data=prov.shp,
               weights=fert.fgls$weights)
summary(fert.wlm)
```

## Map the regression residual

```{r residmap, fig.height=8, fig.width=12}
## Plot Regression Residuals (Bi-polar)
fertResid <- weighted.residuals(fert.wlm)                # Update residuals
hist(fertResid, main="Residuals of Weighted Model")      # Explore distribution to
length(fertResid[fertResid < 0])                         # identify number of pos/neg classes
length(fertResid[fertResid >= 0])

plot(neig.shp,axes=T,col=grey(0.9),border="white",
     xlim=prov.bbox[1,],ylim=prov.bbox[2,])               # first background
mapBiPolar(fertResid, prov.shp,                           # second regression residuals
            neg.breaks=5, pos.breaks=4, break.value=0.0, 
            map.title="Fertility Model Residuals",
            legend.title="Residuals", 
            legend.cex=1.7, add.to.map=T)
```

The map pattern indicates that there are spatial clusters with positive and negative regression residuals. Thus the independence assumption is violated.

## Identify the linkage structure

```{r linkmap, fig.height=8, fig.width=12}
## Plot Augmented Spatial Links among Italian Provinces
## Notes: Shape file has been edited so satellite islands are connected to mainland
## Alternatively spdep::edit.nb function (does not work with RStudio)
prov.link <- poly2nb(prov.shp, queen=F)                          # Generate neighbors links

plot(neig.shp,axes=T,col=grey(0.9),border="white",
     xlim=prov.bbox[1,],ylim=prov.bbox[2,])                      # First background
plot(prov.shp,col="palegreen3" ,border=grey(0.9), axes=T, add=T) # Second plot areas
plot(prov.link,coords=prov.centroid, pch=19, cex=0.1,            # Third plot links focused at centroids
     col="blue", add=T)
title("Augmented Spatial Links among Provinces")                 # Forth add title
box()                                                            # Fifth refresh frame
```

## Perform a test for spatial autocorrelation in the residuals

```{r morantest, fig.height=8, fig.width=12}
prov.linkW <- nb2listw(prov.link, style="W")                   # generated row-sum standardized neighbors list
spOutliers <- moran.plot(weighted.residuals(fert.wlm),         # Moran plot with outlier diagnositics
                         prov.linkW, labels=prov.shp$PROVNAME)          
lm.morantest(fert.wlm, prov.linkW)                             # Test with W-coding scheme
```

Also look up `help(lm.morantest)`. Internally weighted regression residuals are used.

## Maximum likelihood estimation of SAR
Check the online help for the `spdep::spautolm( )` function. Note the use of the `weights` option here. When the `weights` option is used `spautolm` requires a full specification of the model rather than just using the `fert.wlm` object.
```{r SARmodel, message=FALSE, warning=FALSE}
fert.SAR <- spautolm(TOTFERTRAT~ILLITERRAT+FEMMARAGE+DIVORCERAT+TELEPERFAM, 
                     data=prov.shp, weights = fert.fgls$weights,
                     na.action="na.omit", listw=prov.linkW, family="SAR")
summary(fert.SAR)
```

The likelihood ratio test of the SAR model is identically to the hand calculated one. Important here is a slight difference in the calculation when heteroscedasticity is incorporated.
```{r SARlikeratio}
## Likelihood Ratio test (identical to LR from spautolm)
#(likeH0 <- logLik(fert.lm))                     # Use for unweighted model
likeH0 <- fert.fgls$logLikeH1
(likeH1 <- logLik(fert.SAR))

cat("chi-square value:  ", chi <- -2*(likeH0[1]-likeH1[1]))
cat("\nerror-probability: ", pchisq(chi, df=1, lower.tail=F))
```

Check whether the residuals are as expected free of spatial autocorrelation. Because the distribution of ML residuals in unknown, Morans's $I$ is evaluated here using a randomization approach.
```{r SARresid, fig.height=8, fig.width=12}
## Moran test applying randomization because ML may not be normal distributed
plot(neig.shp,axes=T,col=grey(0.9),border="white",
     xlim=prov.bbox[1,],ylim=prov.bbox[2,])                        # first background
mapBiPolar(residuals(fert.SAR), prov.shp,                          # second regression residuals
            neg.breaks=5, pos.breaks=4, break.value=0.0, 
            map.title="Fertility SAR Model Residuals",
            legend.title="Residuals", legend.cex = 1.6,
            add.to.map=T)

## Evaluate ML residuals for spatial autocorrelation
moran.mc(residuals(fert.SAR), prov.linkW, nsim=9999) 

```