---
title: "Maximum likelihood"
author: "Yalin Yang"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: 
          collapsed: false
    number_sections: true          
    toc_depth: 3  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, prompt = TRUE, comment = "R>", cache = FALSE)
```

# Evaluate the likelihood and log-likelhood functions

The following code-chunks evaluate the likelihood function $L(\pi \mid x,n)=\pi^x \cdot (1-\pi)^{n-x}$ and the log-likelihood function of a binary distributed outcomes $x_i \in \{0,1\}$. 

Set up the model:
```{r def}
pi <- 1:99/100                                 # sequence of valid pi values
x <- 7                                         # number of successes
n <- 10                                        # number of trails
likFunc <- function(x,n,pi) pi^x*(1-pi)^(n-x)  # likelihood function
```

It shows, by example, that the maximum $\pi_{max}$ of both functions with the number of successes `r x` and the number of trails `r n` is the *identical*.

# Plot the likelihood function $L(\pi \mid x,n)$
```{r lik}
plot(pi,likFunc(x,n,pi), type="l", xlab=expression(pi), 
     ylab="likelihood function")
abline(v=x/n, lty=5, col="red")
```

# Plot the log-likelihood function $\log(L(\pi \mid x,n))$
```{r loglik}
plot(pi,log(likFunc(x,n,pi)), type="l", xlab=expression(pi), 
     ylab="log-likelihood function")
abline(v=x/n, lty=5, col="red")
```

# The numerical solution for $\hat{\pi} = \frac{x}{n}$
The log-likelihood function is:
$$
\log(L(\pi)) \stackrel{\rm def}{=}
    \log(\pi^x \cdot (1-\pi)^{n-x}) = x \cdot \log \pi + (n-x) \cdot \log (1-\pi)  
$$
Its first derivative becomes:
$$
\frac{d \log(L(\pi))}{d \pi}  =  \frac{x}{\pi} + \frac{n-x}{1-\pi} \cdot -1 =  \frac{x}{\pi} - \frac{n-x}{1-\pi}
$$
Setting the first derivative equal to zero gives the desired estimator \hat{\pi} = \frac{x}{n}:
$$
\frac{x}{\pi} - \frac{n-x}{1-\pi} = 0 \Rightarrow \hat{\pi} = \frac{x}{n}
$$
