---
title: 'Script: MultiWeightedMaxLike'
author: "Michael Tiefelsdorf"
date: "3/29/2020"
output: 
  html_document:
    toc: true
    toc_float: 
          collapsed: false
    number_sections: true          
    toc_depth: 3  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, prompt = TRUE, comment = "R>", cache = FALSE)
rm(list=ls())
library(TexMix)
```

# Read data & explore data
```{r readdata}
setwd("G:\\UTD_Classes\\2020Spring\\GISC7310_AdvancedDataAnalysis\\LC6MLandGLS")
Bladder <- foreign::read.spss("bladder_wmp2.sav",to.data.frame=TRUE)

car::scatterplotMatrix(~bladp2+lungp1+popden, data=Bladder)
car::scatterplotMatrix(~log(bladp2)+log(lungp1)+log(popden), data=Bladder)
hist(Bladder$pop, breaks=20, freq=FALSE, xlab="Population at Risk")
lines(density(Bladder$pop), col="red")
```

Note: these data are _areal aggregates_ therefore there will be some _ecological biases_ in the model estimates.

# Variables description
* The dependent variable `bladp2` measure the age-adjusted male bladder cancer mortality rate per 100.000 population at risk for the 508 US State Economic Areas (SEA) for the period of 1970 to 1994.

* The independent variable `lungp1` is the male lung cancer rate for the period 1950 to 1969. It is a proxy variable for smoking. Smoking is both a risk factor for lung and bladder cancer. Lung cancer has a shorter latency period than bladder cancer. Therefore, the 1950 to 1969 period is used.

* The independent variable `popden` is a proxy variable for behavioral and environmental factors. The populations behavior as well as environmental conditions are different in urban, suburban and rural SEAs.

* The weights variable `pop` is the count of the population at risk in each SEA for the period 1970 to 1994. It varies substantially from SEA to SEA. Thus the SEA's rates of `bladp2` have different denominators and are expected to have varying variances, which cause heteroscedasticity.

# Basic model and heteroscedasticity test

```{r heterotest}
lmBase <- lm(log(bladp2)~log(lungp1)+log(popden), data=Bladder)
summary(lmBase)

car::ncvTest(lmBase, data=Bladder) 
car::ncvTest(lmBase, var.formula=~log(pop), data=Bladder)
```

Concurring with epidemiological knowledge, the higher the smoking rate (i.e., `lungp1`) the higher the bladder cancer rate. More densely populated SEA's have are populated the higher the bladder cancer rate. 

The Breusch-Pagan test `car::ncvTest` preforms a score test of the hypothesis of constant error variance against the alternatives [a] that the error variance changes with the level of the fitted values (first test), or [b] with a linear combination of predictors (second test). Notice that the weights variable in the second test is entered log-transformed.

* The unspecific test without an explicit heteroscedasticity model is borderline significant. 

* In contrast, assuming that the error variance is a function of the underlying population at risk is highly significant.

# Underlying idea of model calibration
See Kleiber & Zeileis pp 76-78 "Weighted Least Squares".  At the first step, use the model 

$$
 \log{\hat{\sigma}_i^2}={\tt fitted}(\log{\epsilon_i^2} \sim \log pop_i) 
$$
to predict the heteroscedastic error variance. In a subsequent step use the inverse variance $w_i=\frac{1}{\exp(\hat{\sigma}_i^2)}$ as weight in an updated regression $y \sim x_1+x_2+ \cdots +x_k, {\tt weight}=w_i$. Note: This way observations with higher uncertainty are down-weighted. It is possible to iterate.

```{r auxreg}
auxreg <- lm(log(residuals(lmBase)^2)~log(Bladder$pop)) 
summary(auxreg)

plot(log(residuals(lmBase)^2)~log(Bladder$pop)); abline(auxreg, col="red")
title("Heteroscedastic lm-Residuals")

## Weighted Regression
predLogSigma2 <- fitted(auxreg)
lmUpdated <- update(lmBase, weights=1/exp(predLogSigma2))
summary(lmUpdated)

```

# Use `lmHetero()` to model Heteroscedasticity
To account for heteroscedasticity related to `log(pop)` the `lmHetero` function in the package `TexMix` can be used. See its online help on how to call the function and what objects it returns.

```{r simpleHetero}
lmH1 <- lmHetero(lnbladd~lnlung1+lnpopden | log(pop), data=Bladder)
summary(lmH1)                         # coefficients of lmH1 and lmBase are similar
names(lmH1)
```

Compared to the OLS model the estimated regression coefficients have not changed substantially because even under heteroscedasticity the OLS coefficients are unbiased. However, their standard errors will change.

The `gamma coefficient` of `log(pop)` is _negative_ and significant. Thus with an increasing population at risk $n_i$, as suggested by theory, the error variance $\sigma_i^2$ shrinks:
$$
\sigma_i^2 \sim \frac{1}{n_i}
$$
The likelihood ratio test indicates that, compared to the log-likelihood of the plain OLS model, the log-likelihood of the adjusted model is _significantly_ larger.

The `lmHetero( )` returns also a vector of case weights, which can be use in the standard `lm( )` function. This allows to perform the standard model diagnostics of the OLS model. The estimated regression coefficients are identical to those of the `lmHetero` model. The residuals obtained with the function `weighted.residuals( )` are free of heteroscedasticity with respect to the weights variable.

```{r weightReg}
## Once weights are estimated, diagnositcs could proceed with lm() and the weights-option
wlm1 <- lm(log(bladp2)~log(lungp1)+log(popden), data=Bladder, weights=lmH1$weights)
summary(wlm1)                         # Results of lmH1 and wlm1 are identical
wResid <- weighted.residuals(wlm1)    # Function to adjust residuals for heteroscedasticity
```

The example shows how a linear combination of several weights variables can be used.
```{r squareHetero}
##
## Model with quadratic weights structure
##
lmH2 <- lmHetero(log(bladp2)~log(lungp1)+log(popden) | log(pop)+I(log(pop)^2), data=Bladder)
summary(lmH2)

##
## Manual LR-Test: Quadratic versus linear weights structue
##
(chi <- -2*(lmH1$logLikeH1-lmH2$logLikeH1))
(pchisq(chi,df=1, lower.tail=F))
```

The manual likelihood ratio test compares the model with `| log(pop)` against the augmented model `| log(pop)+I(log(pop)^2`. In this case, there is no significant improvement using the augmented model and its `Gamma Coefficients` are no longer significant.