---
title: "Bivariate Regression Analysis"
author: "Yalin Yang"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
--- 

# Quantile-Normal Plot

## Initialize data

```{r}
## Data vector with ties in the last four elements
x <- c(4.0,4.4,3.8,2.5,5.1,4.5,3.8,4.8,4.4,4.1)

## Sorting
( xSort <- sort(x) )   # works only on vectors
```

Re-ordering works on matrices and data-frames

```{r}
shuffle <- order(x)      # Order generates a shuffle index
xOrdered <- x[shuffle]   # Shuffle data positions in vector
(cbind(x, shuffle, xOrdered))
```
```{r}
## Ranking data
(xRank <- rank(x, ties.method="random"))      # Explore other methods
```

## Quantiles

```{r}
## quantiles Q[i](p) = (1 - z)*x[j] + z*x[j+1] with 0 <= z <= 1
( quantile(xOrdered,prob=seq(0.1,0.9,by=0.1)) ) 
( quantile(xOrdered,prob=c(0.25,0.5,0.75)) )    # Quartiles
```

## Quantile-Normal Plot

Percentage (probability) points - for tied data use the larger percentile

```{r}
xPercent <- cbind("X-value"=xOrdered, "Percentile a=0.0"=ppoints(xOrdered,a=0.0),
                  "Percentile a=0.5"=ppoints(xOrdered,a=0.5), 
                  "Percentile a=1.0"=ppoints(xOrdered,a=1.0))
round(xPercent, 2)
```

Empirical Distribution Function

```{r fig.height=8, fig.width=12}
summary(ecdf(x))
plot(ecdf(x))
```

QQ plot

```{r fig.height=8, fig.width=12, message=FALSE, warning=FALSE}
car::qqPlot(x)
```

# Box Cox Transformation

## Initialize data

```{r}
library(car)
setwd("G:\\UTD_Classes\\2020Spring\\GISC7310_AdvancedDataAnalysis\\02Bivariate Regression Analysis")
Concord <- foreign::read.spss("Concord1.sav",to.data.frame=TRUE)
reg01 <- lm(water81~income, data=Concord)
summary(reg01)
```

## Residual Analysis

```{r fig.height=8, fig.width=12}
hist(resid(reg01))
```

```{r}
round(sum(resid(reg01)),14)
cbind("Coef"=coef(reg01), confint(reg01, level=0.95))
```

Prediction and Observation

### Line confidence interval & fit 

```{r}
predDf <- data.frame(income=min(Concord$income):
          max(Concord$income))   # data-frame for independent vars
(predDf <- data.frame(predDf, predict(reg01, 
      newdata=predDf, interval="confidence", level=0.95)))[1:10,] # Line confidence interval & fit
```

```{r fig.height=8, fig.width=12}
plot(water81~income,data=Concord)
lines(predDf$income,predDf$fit,col="red")  # predicted value
lines(predDf$income,predDf$lwr,col="green") # lower confidence interval limits
lines(predDf$income,predDf$upr,col="green") # upper confidence interval limits
abline(h=mean(Concord$water81),v=mean(Concord$income),lty=3)  # Regression line goes thru the means
```

### Point confidence interval & fit

```{r}
predDf <- data.frame(income=min(Concord$income):
                       max(Concord$income))   # data-frame for independent vars
(predDf <- data.frame(predDf, predict(reg01, 
       newdata=predDf, interval="prediction", level=0.95)))[1:10,]  # Point confidence interval & fit
```

```{r fig.height=8, fig.width=12}
plot(water81~income,data=Concord)
lines(predDf$income,predDf$fit,col="red")   # predicted value
lines(predDf$income,predDf$lwr,col="green") # lower confidence interval limits
lines(predDf$income,predDf$upr,col="green") # upper confidence interval limits
abline(h=mean(Concord$water81),
       v=mean(Concord$income),lty=3) # Regression line goes thru the means
```

### Scatter Plot

```{r fig.height=8, fig.width=12}
car::scatterplot(water81~income, data=Concord, 
                 main="Concord Households: Water Consumption against Income" )
```

**The residual variance is not constant and mean are not equal to 0**

```{r fig.height=8, fig.width=12}
plot(resid(reg01)~fitted(reg01))
abline(h=0,col= "red")
```

# Box Cox Transformation

## Initialize data

```{r}
setwd("G:\\UTD_Classes\\2020Spring\\GISC7310_AdvancedDataAnalysis\\02Bivariate Regression Analysis")  
library(foreign); library(car)
myPower <- read.spss("DallasTempPower.sav", to.data.frame= TRUE)

myPower$powPerDay <- myPower$kWhBill/myPower$DaysBill # calculate kWh per day

## Exploration
summary(myPower)
```

```{r fig.height=8, fig.width=12}
boxplot(powPerDay~Month, data=myPower)
abline(h=mean(myPower$powPerDay, na.rm=TRUE), lty=5, col="red")
```

```{r fig.height=8, fig.width=12}
boxplot(myPower[, c("MinTemp","AveTemp","MaxTemp")])
```

```{r fig.height=8, fig.width=12}
hist(myPower$AveTemp, breaks=seq(40,100, by=5))
```

## Box Cox Transformation

### Check normality

```{r}
e1071::skewness(myPower$powPerDay, na.rm=TRUE)
```
```{r fig.height=8, fig.width=12}
car::qqPlot(myPower$powPerDay)
```

**Two methods for testing**

```{r}
shapiro.test(myPower$powPerDay)
```

```{r}
ks.test(myPower$powPerDay, pnorm,                   # the ks test has not as much power
        mean=mean(myPower$powPerDay), sd=sd(myPower$powPerDay))
```

### Find Box-Cox lambda

Explore different lambda parameters

```{r fig.height=8, fig.width=12}
symbox(~powPerDay, data=myPower)
```

Test indicates log-transformation sufficient

```{r}
summary(powerTransform(lm(powPerDay~1, data=myPower)))
```

```{r message=FALSE, warning=FALSE}
lambda <- powerTransform(lm(powPerDay~1, data=myPower))$lambda
myPower$bc.powPerDay <- car::bcPower(myPower$powPerDay, lambda=lambda)
e1071::skewness(myPower$bc.powPerDay)
```
```{r fig.height=8, fig.width=12}
car::qqPlot(myPower$bc.powPerDay)
```
```{r}
shapiro.test(myPower$bc.powPerDay) 
```

### Scatterplot with loess smoother

```{r fig.height=8, fig.width=12}
scatterplot(powPerDay~AveTemp, data=myPower)
```

## Simultaneously transform a set of variables 
```{r}
summary(lambda <- powerTransform(lm(cbind(powPerDay,AveTemp)~1, data=myPower)))
myPower <- data.frame(myPower,bcPower(cbind(myPower$powPerDay,myPower$AveTemp), 
                                      coef(lambda, round=T))) # add transformed variables to myPower
```

# Box Cox Transformation [Negative Values]

## Z-Gamma Transformation

```{r fig.height=8, fig.width=12}
## Example of the z-Gamma transformation
zGamma <- function(x, gamma){(x+sqrt(x^2+gamma^2))/2}
x <- seq(-2,4, by=0.1)
gamma <- 3
zx <- zGamma(x, gamma)
plot(x,zx, type="l", xlab="X with negative values", ylab="Transformed z(x)",
     main="Box-Cox Family with Negative Values and Gamma=1")
```

## An positvely skewed distribution with small negative value

```{r fig.height=8, fig.width=12}
x <- rbeta(100, shape1=2, shape2=5)-0.1
hist(x, main="Beta distribution with rbeta(100, shape1=2, shape2=5)-0.1")
```

### Use powerTransform

```{r}
summary(lambda <- powerTransform(x~1, family="bcnPower"))
```

### Z-Gamma Transformation

```{r fig.height=8, fig.width=12}
x <- sort(x)
zx <- zGamma(x,coef(lambda)[2])
plot(x,zx, type="l", xlab="X with negative values", ylab="Transformed z(x)",
     main="Box-Cox Family with Negative Values and Gamma=0.713883")
```

### Box Cox Transformation

```{r fig.height=8, fig.width=12}
x.bcn <- bcnPower(x, lambda=coef(lambda)[1], gamma=coef(lambda)[2])
hist(x.bcn, main="Box-Cox transformation with Negative Values")
```

