---
title: "Bivariate Regression Analysis"
author: "Yalin Yang"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
--- 

# Scatter Plot & Effect Plots

## Initialize Data

**Relevant variables:**

* price:   median home price in community
* crime:   crime rate
* nox:     nitrogen oxide in the air
* dist:    weighted distance to five employment centers
* rooms:   average number of in houses in the community
* stratio: Student-teacher ratio of schools in the community
* proptax: property tax in community per $1000 home value

```{r message=FALSE, warning=FALSE}
library(car); library(effects)
hprice2 <- foreign::read.dta("http://fmwww.bc.edu/ec-p/data/wooldridge/hprice2.dta")
summary(hprice2)
```

## Check normality

```{r}
summary(powerTransform(cbind(price,crime,nox,dist,rooms,stratio)~1, data=hprice2))
```

## Scatter Plot

```{r fig.height=10, fig.width=15, message=FALSE, warning=FALSE}
scatterplotMatrix(~price+crime+nox+dist+rooms+stratio, data=hprice2, pch=1, 
                  smooth=list(span = 0.35,lty.smooth=1, col.smooth="red", col.var="red"),
                  regLine=list(col="green"))
```

**Applied log-transformation on highly positive skewed variables**

```{r fig.height=10, fig.width=15}
scatterplotMatrix(~log(price)+log(crime)+log(nox)+log(dist)+rooms+stratio, 
                  data=hprice2, pch=1, 
                  smooth=list(span = 0.35,lty.smooth=1, col.smooth="red", col.var="red"),
                  regLine=list(col="green"))
```

## Regression modeling on untransformed data

### Plain Model

```{r}
mod0 <- lm(price~nox+dist+rooms+stratio, data=hprice2)
summary(mod0)
```

### Quadratic term (Could use turkey test)

```{r}
mod0 <- lm(price~nox+dist+rooms+I(rooms^2)+stratio, data=hprice2)
summary(mod0)
```

### Effect Plots

```{r fig.height=8, fig.width=12}
mod0.eff <- allEffects(mod0, xlevels=list(rooms=3:9))
plot(mod0.eff, "rooms", main="Non-linear effect of # of rooms")
```

## Regression modeling on log-form data

### Base model

```{r}
mod1 <- lm(log(price)~log(crime)+log(nox)+log(dist)+rooms+stratio, data=hprice2)
summary(mod1)
```

### Quadratic effects

```{r}
mod2 <- lm(log(price)~log(crime)+log(nox)+log(dist)+rooms+I(rooms^2)+stratio, data=hprice2)
summary(mod2)
```

**anova test**
```{r}
anova(mod1,mod2)
```

### Explore non-linear effect

Turning point:  $-\frac{b}{2a}$  => $0.672/(2*0.072)=4.667$


```{r fig.height=8, fig.width=12}
mod2a.eff <- allEffects(mod2, xlevels=list(rooms=3:9))
plot(mod2a.eff, "rooms", ylab="log(price)", main="Non-linear effect of # of rooms")
```

### Transfer dependent variable to the original scale

```{r fig.height=8, fig.width=12}
mod2b.eff <- allEffects(mod2, xlevels=list(rooms=3:9),
                       transformation=list(link=log, inverse=exp))
plot(mod2b.eff, "rooms", ylab="price", main="Non-linear effect of # of rooms")
```

# Conditional Effect Plot

## Initialize Data

**Key variables:**
* stndfnl: 	Standardized outcome on final exam
* atndrte: 	Percentage of class attendence
* priGPA:  	Prior college grade point average
* ACT:     	American College Testing score

```{r}
library(car); library(effects)
attend <- foreign::read.dta("http://fmwww.bc.edu/ec-p/data/wooldridge/attend.dta")
summary(attend)
```

## Regression Modeling

**Scatter Plot**

```{r fig.height=10, fig.width=15}
scatterplotMatrix(~stndfnl+atndrte+priGPA+ACT, data=attend, pch=1, 
                  smooth=list(span = 0.35,lty.smooth=1, col.smooth="red", col.var="red"),
                  regLine=list(col="green"))
```

**Base model**

```{r}
mod1 <- lm(stndfnl~atndrte+priGPA+ACT, data=attend)
summary(mod1)
```

**With interaction. Notice the "*" in the formula**

```{r}
mod2 <- lm(stndfnl~atndrte*priGPA+ACT, data=attend)
summary(mod2)
```

**Partial F-test**
here equal to the t-test for the interaction term

```{r}
anova(mod1,mod2)
```

## Conditional Effect Plot

Effect of attendence rate at **average priGPA**

```{r}
(b <- coef(mod2))
(mean(attend$priGPA))
```

```{r}
cat("Partial effect of atndrte for priGPA=2.59:", b["atndrte"]+mean(attend$priGPA)*b["atndrte:priGPA"])
```

Test partial effect at priGPA=mean(attend$priGPA)

```{r}
linearHypothesis(mod2, c("atndrte+2.59*atndrte:priGPA"))
```

```{r fig.height=8, fig.width=12}
plot(allEffects(mod2))
```

```{r fig.height=8, fig.width=12}
plot(Effect(c("atndrte","priGPA"), mod2, 
            xlevels=list(priGPA=seq(1.5, 3.5, by=0.5))))
```

# Partial Effects Demo

## Initialize Data
```{r fig.height=8, fig.width=12}
concord <- foreign::read.spss("Concord1.sav",to.data.frame=T)
car::scatterplot(water81~income, data=concord, 
                 xlab="Income in $1,000",ylab="Wate consumption in 1981 in ft^2")
```

## Demonstration: Partial Effects

```{r}
mod1.lm <- lm(water81 ~ income, data=concord)
summary(mod1.lm)
```

```{r}
mod2.lm <- lm(water81 ~ income + water80, data=concord)
summary(mod2.lm)
```

### Controling for water80

```{r fig.height=6, fig.width=10}
y.x2 <- residuals(lm(water81 ~ water80, data=concord)) 
plot(concord$water81,y.x2)
```
```{r}
cor(concord$water81,y.x2)
cor(concord$water80,y.x2)
```

```{r}
x1.x2 <- residuals(lm(income ~ water80, data=concord))
cat("Sum of residuals for Water81|Water80:",sum(y.x2),"\n")                   #sum of residuals = 0
cat("Sum of residuals for Income|Water80: ",sum(x1.x2),"\n")
```

```{r fig.height=8, fig.width=12}
scatterplot(y.x2~x1.x2, xlab=bquote(hat(bold(e))["Income|Water80"]),
            ylab=bquote(hat(bold(e))["Water81|Water80"]))   
```

**compare with coefficients of mod2.lm**

```{r}
summary(lm(y.x2 ~ x1.x2-1))
```

### Controling for income

```{r fig.height=8, fig.width=12}
y.x1 <- residuals(lm(water81 ~ income, data=concord))                       
x2.x1 <- residuals(lm(water80 ~ income, data=concord))
scatterplot(y.x1~x2.x1, xlab=bquote(hat(bold(e))["Water81|Income"]),          
            ylab=bquote(hat(bold(e))["Water80|Income"]))   
```

**compare with coefficients of mod2.lm**

```{r}
summary(lm(y.x1 ~ x2.x1-1)) 
```

# Beta Coefficients

## Linear Modeling

full model
```{r}
mod3.lm <- lm(water81 ~ income+water80+educat+retire+peop81+cpeop, data=concord)
summary(mod3.lm)
```


convert factor to metric. retired:"yes"=1,"no"=0 

```{r}
concord$retireDummy <- as.numeric(concord$retire)-1
```

Transfer all variables to scaled form

```{r}
concordNew <- concord[ ,sapply(concord,is.numeric)]  # Remove non-numeric variables
concordScale <- as.data.frame(scale(concordNew))     # apply z-transformation with scale function
                                                     # dataframe concordScale holds the transformed values
mod4.lm <- lm(water81 ~ -1+income+water80+educat+retireDummy+peop81+cpeop, data=concordScale)
summary(mod4.lm) 
```

## coefficent plot

coefficent plot useful for beta weights because the parameters are on the same scale

```{r fig.height=8, fig.width=12}
library (coefplot)
coefplot(mod4.lm)
```

## Partial F-test

H0: $\beta_{RETIRE}=\beta_{CPEOP}=0$

```{r}
mod5.lm <- lm(water81 ~ water80+income+educat+peop81, data=concord)
summary(mod5.lm)
```

**Compare both models**

```{r}
anova(mod5.lm,mod3.lm)
```

**Nested F test**
Equal to t test
```{r}
Anova(mod3.lm)
```

## Stepwise Model

```{r}
null.lm <- lm(water81~1, data=concord)
mod6.step <- step(null.lm,
                  scope= ~income+water80+educat+retire+peop81+cpeop,
                  direction="forward")
summary(mod6.step)
```

Alternative stepwise specification

```{r}
mod7.step <- step(mod3.lm, scope=list(lower=null.lm, upper=mod3.lm), direction="backward")
summary(mod7.step)
```

## Conditional Effects

```{r}
concord$clogpeop <- log(concord$peop81 / concord$peop80)
mod08.lm <- lm(log(water81)~log(income)+log(water80)+educat+retire+log(peop81)+clogpeop, data=concord)
summary(mod08.lm)
```

### All effects at mean level of remaing variavles

```{r fig.height=8, fig.width=12}
plot(allEffects(mod08.lm, transformation=list(link=log, inverse=exp)), ylab="water81")
```

### get value ranges

```{r}
summary(concord)
summary(log(concord$water80))
summary(log(concord$peop81))
```

### Income effect for a low consumer profile

```{r fig.height=8, fig.width=12}
plot(effect("log(income)", mod08.lm, given.values=c("log(water80)"=7.3, "educat"=18, "log(peop81)"=0.7),
            transformation=list(link=log, inverse=exp) ), 
     ylim=c(log(1000),log(6000)), ylab="water81", 
     main="Income Effect of Low Water Consumer")
```

### Income effect for an average consumer profile

```{r fig.height=8, fig.width=12}
plot(effect("log(income)", mod08.lm, given.values=c("log(water80)"=7.7, "educat"=14, "log(peop81)"=1.0),
            transformation=list(link=log, inverse=exp) ), 
     ylim=c(log(1000),log(5000)), ylab="water81", 
     main="Income Effect of Average Water Consumer")
```

### Income effect for a high consumer profile

```{r fig.height=8, fig.width=12}
plot(effect("log(income)", mod08.lm, given.values=c("log(water80)"=8.3, "educat"=10, "log(peop81)"=1.4),
            transformation=list(link=log, inverse=exp) ), 
     ylim=c(log(1000),log(5000)), ylab="water81", main="Income Effect of High Water Consumer")
```

# Factor Variable Analysis

## Initialize Data

```{r}
wells <- foreign::read.spss("wells.sav",to.data.frame=TRUE)
summary(wells)
```

## Explore Coding Scheme of Factors

See coding of factor

```{r}
class(wells$deep)
contrasts(wells$deep)
```

Change to 1,0,-1 coding

```{r}
contrasts(wells$deep) <- "contr.sum" 
contrasts(wells$deep)
```

Change back to 0,1 coding

```{r}
contrasts(wells$deep) <- "contr.treatment"
contrasts(wells$deep)
```

## Prepare data for analysis

```{r}
wells1 <- na.omit(wells)                               # drop observation 18 with missing values for Chlor
wells1$logChlor <- log(wells1$chlor)                   # Transform to natural logarithm
wells1$logDist <- log(wells1$droad)                    # Transform to natural logarithm
(wells1$deepdum <- as.numeric(unclass(wells1$deep))-1)   # Convert factor to numeric 0/1 dummy
attach(wells1)
```

```{r fig.height=8, fig.width=12}
scatterplot(logChlor~logDist|deep, smooth=F, data=wells1)
```

Symbols & colors for well type

```{r fig.height=8, fig.width=12}
wellSymbol <- ifelse(deep==levels(deep)[1],15,16)
wellCol <- ifelse(deep==levels(deep)[1],"red","blue")
plot(droad,chlor, log="xy", pch=wellSymbol, col=wellCol)
legend("topright",legend=c("shallow","deep"), 
       title="Well Type:", col=c("red","blue"),pch=c(15,16))
```

## Linear Model for factor variables

**No intercept Model**

```{r}
mod0 <- lm(logChlor ~ deep -1) # One way analysis of variance
summary(mod0)             # Suppressing intercept gives mean levels
```

**intercept model**

```{r}
mod1 <- lm(logChlor ~ deep)   # One-way analysis of variance
summary(mod1)
```

### Standard Regression

```{r}
mod2 <- lm(logChlor ~ logDist)
summary(mod2)
```

```{r fig.height=8, fig.width=12}
plot(logDist,logChlor,pch=wellSymbol,col=wellCol)      # just distance not interaction
abline(mod2)
legend("topright",legend=c("shallow","deep"), title="Well Type", col=c("red","blue"),pch=c(15,16))
```

### Regression with intercept dummy

```{r}
mod3 <- lm(logChlor ~ deep + logDist)
summary(mod3)
```

```{r fig.height=8, fig.width=12}
plot(logDist,logChlor,pch=wellSymbol,col=wellCol)      # intercept dummy
abline(mod3$coef[1],mod3$coef[3],col="blue")
abline(mod3$coef[1]+mod3$coef[2],mod3$coef[3],col="red")
legend("topright",legend=c("shallow","deep"), title="Well Type", col=c("red","blue"),pch=c(15,16))
```

### Regression with slope dummy

```{r}
mod4 <- lm(logChlor ~ logDist + logDist:deep) 
summary(mod4)
```

```{r fig.height=8, fig.width=12}
plot(logDist,logChlor,pch=wellSymbol,col=wellCol)      # slope dummy
abline(mod4$coef[1],mod4$coef[2],col="red")
abline(mod4$coef[1],mod4$coef[2]+mod4$coef[3],col="blue")
legend("topright",legend=c("shallow","deep"), title="Well Type", col=c("red","blue"),pch=c(15,16))
```

### Regression with intercept and slope dummy

```{r}
mod5 <- lm(logChlor ~ deep*logDist)
summary(mod5)
```

```{r fig.height=8, fig.width=12}
mod5 <- lm(logChlor ~ deep + logDist + logDist:deep) 
plot(logDist,logChlor,pch=wellSymbol,col=wellCol)      # intercept and slope dummy
abline(mod5$coef[1],mod5$coef[3],col="red")
abline(mod5$coef[1]+mod5$coef[2],mod5$coef[3]+mod5$coef[4],col="blue")
legend("topright",legend=c("shallow","deep"), title="Well Type", col=c("red","blue"),pch=c(15,16))
```

```{r}
anova(mod2,mod5)
```

