---
title: "Instrumental Variable Regression"
author: "Yalin Yang"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
    toc_depth: '3'
--- 

# Instrumental Variable Regression

## Quickview of Dataset

```{r message=FALSE, warning=FALSE}
library(AER)
mroz <- foreign::read.dta("http://fmwww.bc.edu/ec-p/data/wooldridge/mroz.dta")
## Drop records without wage information
mrozwage <- subset(mroz, !is.na(wage))
```

OLS slope $\beta_1 = Cov(y,x) / Var(x)$

```{r}
with(mrozwage, cov(log(wage),educ) / var(educ))
```

OLS with linear mode

```{r}
summary(lm(log(wage)~educ, data=mrozwage))
```

The correlation between the disturbuance and endogenous variable (One type of Heteroscedasticity)

In this situation,we focous on the relationship between disturbuance and independent(endogenous) variable not dependent variable

```{r}

cor(log(mrozwage$wage) - mean(log(mrozwage$wage)), mrozwage$educ)

```

## Instrumental regression

```{r}
summary(ivreg(log(wage)~educ | fatheduc, data=mrozwage))
```

Manually apply 2SLS with augmented model

* educ is an endogenous regressor
* exper is an exogenous regressor
* mother's and father's education are instruments for ability

```{r fig.height=10, fig.width=15, message=FALSE, warning=FALSE}
scatterplotMatrix(~log(wage)+exper+I(exper^2)+motheduc+fatheduc, data=mrozwage,
                  pch=1, smooth=list(span = 0.35,lty.smooth=1, col.smooth="red", col.var="red"),
                  regLine=list(col="green"))
```

### 1st stage

Regression between endogenous and exogenous plus instruments
[endogenous should be independent with exogenous but not instruments]
```{r}
stage1 <- lm(educ~exper+I(exper^2)+motheduc+fatheduc, data=mrozwage)
summary(stage1)
```

### 2nd stage 
With incorrect standard errors

```{r}
preEduc <- fitted(stage1)
# preEduc <- residuals(stage1)
stage2 <-lm(log(wage)~preEduc+exper+I(exper^2), data=mrozwage)
summary(stage2)
```

### Biased OLS estimates

```{r}
summary(lm(log(wage)~educ+exper+I(exper^2), data=mrozwage))
```

### IV Reg model

```{r}
aut.2SLS<-ivreg(log(wage)~educ+exper+I(exper^2) |
                          motheduc+fatheduc+exper+I(exper^2) , data=mrozwage)
summary(aut.2SLS, diagnostics = TRUE)
```

### Partial F-test
Test relevance of instruments (weak instruments) with partial F-test

```{r}
stage1.aux <- lm(educ~exper+I(exper^2), data=mrozwage)
anova(stage1.aux,stage1)
```

### Modified Hausman test
Test educ for endogeneity (see the coefficient for resid(stage1))

```{r}
res.2SLS <- lm(log(wage)~educ+exper+I(exper^2)+resid(stage1), data=mrozwage)
coeftest(res.2SLS)
```

### Sargan test
Test of exogeneity of instruments
```{r}
res.aux <- lm(resid(aut.2SLS) ~ motheduc+fatheduc+exper+I(exper^2) , data=mrozwage)
(r2 <- summary(res.aux)$r.squared)
(n <- nobs(res.aux))
(teststat <- n*r2)
(pval <- 1-pchisq(teststat, df=1))
```

# Chi-Square Test

$\chi^2$ Test is very sensitive to the sample size if the sample size go larger, the significance for the Chi-Square Test would increase substantially (p-value would decrese)

## Small sample size test

```{r}
(lowCount <- matrix(c(6,8,7,10),nrow=2))
```

```{r}
(low.test <- chisq.test(lowCount, correct=F)) 
```
Low expected counts
```{r}
low.test$expected 
```
Get p-value through simulation
```{r}
chisq.test(lowCount, simulate.p.value=T) 
```

## Rescaling the sample size

```{r}
(hiCount <- lowCount*1000)
```

chi-square increased by 1000

```{r}
(hi.test <- chisq.test(hiCount, correct=F))
```

```{r}
hi.test$expected 
```

```{r}
chisq.test(hiCount, simulate.p.value=T) 
```

# Stock & Watson Smoking Dataset Modeling

* Cross-sectional data for 48 contiguous U.S. states from 1985-1995
* packpc: average number of packs sold per capita in a year
* avgprs: average annual expenditure per person
* income: average income/pop in $1000
* taxs: proportion of sales tax on each package
* tax: general sales tax rate. May depend on incomepp

## Quick view of dataset

```{r fig.height=10, fig.width=15}
cig <- foreign::read.dta("http://fmwww.bc.edu/ec-p/data/stockwatson/cig85_95.dta")
cig$incomepp <- cig$income/cig$pop
cig <- cig[cig$year==1995,]

scatterplotMatrix(~packpc+avgprs+incomepp, data=cig,
                  pch=1, smooth=list(span = 0.35,lty.smooth=1, col.smooth="red", col.var="red"),
                  regLine=list(col="green"))
```

**Log transformation**

```{r fig.height=10, fig.width=15}
scatterplotMatrix(~log(packpc)+log(avgprs)+log(incomepp), data=cig,
                  pch=1, smooth=list(span = 0.35,lty.smooth=1, col.smooth="red", col.var="red"),
                  regLine=list(col="green"))
```

## Regression Modeling

Misspecified elasticity model without income

```{r}
cig.lm <- lm(log(packpc)~log(avgprs), data=cig)
summary(cig.lm)
```

Income adjusted elasticity model

```{r}
cig.lm <- lm(log(packpc)~log(avgprs)+log(incomepp), data=cig)
summary(cig.lm)
```

## IV Regression
Evaluate first stage of IV regression
```{r fig.height=8, fig.width=12}
scatterplotMatrix(~log(avgprs)+tax+taxs, data=cig,
         pch=1, smooth=list(span = 0.35,lty.smooth=1, col.smooth="red", col.var="red"),
                  regLine=list(col="green"))
```

```{r}
cig.rf <- lm(log(avgprs)~tax+taxs+log(incomepp), data=cig)
summary(cig.rf)
```

2nd estimation

```{r}
cig.iv <-ivreg(log(packpc)~log(avgprs)+log(incomepp)|
                     tax+taxs+log(incomepp), data=cig)
summary(cig.iv, diagnostics=T)
```

